Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-09-07 manuscript/manuscript.Rmd
title and terminology update.

---
title: "Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
abstract: |
  ```{=latex}
  This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product. Here I propose the following three basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.

@@

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology. Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].

As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022]. Unlike controlled experiments, variability could be easily observed in actualistic experiments by design. This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epdedimiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007]. Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681] and the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."

@@

Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts. The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders. By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts [@pargeter2023: 164]. This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions. Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise. It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage. By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].

@@

The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future. Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.

@@

Through the broadening of traditional data types and recording methods revolving around experimental replicas *per se*, the Triple P conceptual framework allows the amplified multiscale expression of material cultural variability. It is also compatible with many theoretical orientations, ranging from behavioral archaeology (emphasis on video recording of behavioral processes) through evolutionary archaeology (emphasis on the amplification of variability) to post-processual archaeology (emphasis on perception through ethnography). In terms of its research practice, it embraces a collaborative mode of knowledge production by involving a more diverse pool of stakeholders. The innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework has a huge potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.

>>>>

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
abstract: |
  ```{=latex}
  This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product. Here I propose the following three basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.

@@

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology. Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].

As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022]. Unlike controlled experiments, variation could be easily observed in actualistic experiments by design. This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epdedimiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007]. Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681] and the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."

@@

Emphasizing variation at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts. The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders. By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts [@pargeter2023: 164]. This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions. Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise. It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage. By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variation very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].

@@

The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future. Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.

@@

Through the broadening of traditional data types and recording methods revolving around experimental replicas *per se*, the Triple P conceptual framework allows the amplified multiscale expression of material cultural variation. It is also compatible with many theoretical orientations, ranging from behavioral archaeology (emphasis on video recording of behavioral processes) through evolutionary archaeology (emphasis on the amplification of variation) to post-processual archaeology (emphasis on perception through ethnography). In terms of its research practice, it embraces a collaborative mode of knowledge production by involving a more diverse pool of stakeholders. The innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework has a huge potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.
